---
title: "The 2000 Florida election"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: css/style.css
runtime: shiny_prerendered
description: >
  Making a choropleth
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(sf)
library(maps)
knitr::opts_chunk$set(message=F, warning=F)
florida_election <- readRDS('www/florida_election.rds')
states <- st_as_sf(maps::map('state', plot = F, 
                             fill = T)) %>% 
  cbind(st_coordinates(st_centroid(.))) %>% 
  mutate(ID = str_to_title(ID))
counties <- st_as_sf(maps::map('county', plot = F, 
                               fill = T), crs=4269)
counties <- counties %>% 
  dplyr::filter(str_detect(ID, 'florida'))
counties <- counties %>% 
  separate(ID, c('State','County'), sep = ',') %>% 
  mutate_at(vars(State:County), str_to_title)
election_by_county <- counties %>% 
  left_join(florida_election)

```
# Introduction

We're going to end up with a choropleth map that shows the percentage of 
votes for Gore by county in Florida. Our final map will look something like this:

```{r, echo=FALSE, fig.height=5}
library(maps)
library(sf)
library(spData)
states <- st_as_sf(maps::map('state', plot = F, 
                             fill = T)) %>% 
  cbind(st_coordinates(st_centroid(.))) %>% 
  mutate(ID = str_to_title(ID))
counties <- st_as_sf(maps::map('county', plot = F, 
                               fill = T), crs=4269)
counties <- counties %>% 
  dplyr::filter(str_detect(ID, 'florida'))
counties <- counties %>% 
  separate(ID, c('State','County'), sep = ',') %>% 
  mutate_at(vars(State:County), str_to_title)
election_by_county <- counties %>% 
  left_join(florida_election)
plt_map <- ggplot(data = world)+
  geom_sf(aes(geometry=geom)) + 
  geom_sf(data = states, aes(geometry=geom), fill = NA) + 
  geom_sf(data = election_by_county,
          aes(fill=Gore_perc, 
              geometry=geom)) + 
  geom_label(data = states %>% 
               dplyr::filter(ID != 'Florida'), #<<
             aes(X, Y, label = ID), #<<
            size = 5) +
  coord_sf(xlim = c(-88, -78), ylim = c(24.5, 33),  
           expand = F) +
  labs(x = '', y = '', fill = 'Percentage for Gore') + 
  scale_fill_viridis_c(option = 'plasma')
plt_map
```

# Creating a map

# First, let's make a map of the United States

```{r f1, eval = T, echo = T}
library(maps)
library(sf)
states <- st_as_sf(maps::map('state', plot = F, 
                             fill = T)) %>% 
  cbind(st_coordinates(st_centroid(.))) %>% 
  mutate(ID = str_to_title(ID))

ggplot(data = states)+
  geom_sf() 
```

# Add labels to the map

```{r map1, exercise=TRUE, exercise.eval=FALSE}
ggplot(data = states)+
  geom_sf() +
  geom_text(data = states, _________, 
            size = 5)  
```

```{r map1-solution}
ggplot(data = states)+
  geom_sf() +
  geom_text(data = states, aes(X, Y, label=ID), 
            size = 5)
```


# And focus in on the southeast

```{r}
ggplot(data = states)+
  geom_sf() +
  geom_text(data = states, aes(X, Y, label=ID), 
            size = 5) +
  coord_sf(xlim = c(-88, -78), ylim = c(24.5, 33), 
           expand = F)

```


# Now we need the county boundaries

We can get them from the **maps** package. Note that we're specifying a CRS value, which defines a particular projection. 

```{r g2, echo = T, exercise=TRUE, exercise.eval=TRUE}
library(maps)
counties <- st_as_sf(maps::map('county', plot = F, 
                               fill = T), crs=4269)
counties <- counties %>% 
  dplyr::filter(str_detect(ID, 'florida'))
counties <- counties %>% 
  separate(ID, c('State','County'), sep = ',') %>% 
  mutate_at(vars(State:County), str_to_title)
```

Play around with the code above and see what `counties` looks like. You'll find that it's  data frame. This means that we'll have a pretty easy job putting the 
election data and geographic data together.


## Merge the election data with the map data


```{r g3, echo = T, exercise=TRUE}
election_by_county <- counties %>% 
  _____join(florida_election)
head(election_by_county, 2)
```

```{r g3-solution}
election_by_county <- counties %>% 
  left_join(florida_election)
head(election_by_county, 2)
```

Figure out which of `inner_join`, `left_join`, or `full_join` you want to use here.

# Now we can plot


```{r g4, echo = T, exercise=TRUE, exercise.eval=FALSE}
ggplot(election_by_county) + 
  ________________________ + 
  scale_fill_viridis_c(option = 'plasma') + 
  labs(caption = 'Source: Wikipedia')
```

```{r g4-solution}
ggplot(election_by_county) + 
  geom_sf(aes(fill = Gore_perc)) + 
  scale_fill_viridis_c(option = 'plasma') + 
  labs(caption = 'Source: Wikipedia')
```


## Add labels

We can clean this up a bit, and add surrounding states.

For this, we make the full US map and limit it to Florida, as before. 
We then add some labels for the names. Note 


```{r g5, eval = T, echo = T, error = T, exercise=TRUE, exercise.eval=TRUE}
plt_map <- ggplot()+
  geom_sf(data = states, fill = NA) + 
  geom_sf(data = election_by_county, 
          aes(fill=Gore_perc)) + 
  geom_label(data = states , #<<
             aes(X, Y, label = ID), #<<
            size = 5) +
  coord_sf(xlim = c(-88, -78), ylim = c(24.5, 33),  
           expand = F) +
  labs(x = '', y = '', fill = 'Percentage for Gore') + 
  scale_fill_viridis_c(option = 'plasma')
plt_map
```

Modify this code so you don't get the label for Flordia printed.

```{r g5-hint-1}
plt_map <- ggplot()+
  geom_sf(data = states, fill = NA) + 
  geom_sf(data = election_by_county, 
          aes(fill=Gore_perc)) + 
  geom_label(data = states %>% #<<
               dplyr::filter(ID != 'Florida'), #<<
             aes(X, Y, label = ID), #<<
            size = 5) +
  coord_sf(xlim = c(-88, -78), ylim = c(24.5, 33),  
           expand = F) +
  labs(x = '', y = '', fill = 'Percentage for Gore') + 
  scale_fill_viridis_c(option = 'plasma')
plt_map
```



And there you have it, Not too bad, right?

The process is 

1. Find the data for the geometries you want to see
1. Find the actual map data you can use (you can always convert it using `st_as_sf`)
1. Do a join to put the data and the map geometries together
1. Plot away. 
