---
title: "Displaying regression results"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: css/style.css
runtime: shiny_prerendered
description: >
  Displaying regression results
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(learnr)
mpg1 <- mpg %>% 
  mutate(year = as.factor(year)) %>% 
  mutate(trans = str_extract(trans, '[:alnum:]+')) %>% 
  mutate(cyl = as.factor(cyl))
fit <- lm(cty ~ year + trans + cyl + drv + class, data=mpg1)
out <- tidy(fit, conf.int=TRUE)
```

# A regression model

We will start with the `mpg` dataset that comes with the **ggplot2** package, 
and fit a linear regression model to see how city fuel efficiency depends on
the year, transmission, number of cylinders, drive and class of vehicle

A lot of the work in this tutorial will use string manipulations using functions
from the **stringr** package. This package does have a [cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/strings.pdf) that you can download.

## Taking a look at the data

```{r}
head(mpg)
```

A few things come to my attention. 

1. The `drv` variable has additional information that we don't need. We just want to keep auto/manual status
1. We only have 2 years in this data, 1999 and 2008. We probably should convert it to a factor
1. I'd like to treat the number of cylinders as a factor, since I'm not sure there will be a linear relationship between number of cylinders and fuel efficiency. I can make it linear later if I need to.

## Data munging

```{r g1, eval=TRUE, exercise=TRUE, exercise.eval=TRUE}
mpg1 <- mpg %>% 
  mutate(year = as.factor(year)) %>% 
  mutate(trans = str_extract(trans, '[:alnum:]+')) %>% 
  mutate(cyl = as.factor(cyl))
```

The one piece you may not know here is `str_extract` and `[:alnum:]`. A good start with this is the _regular expressions_ vignette in the **stringr** package. You can access it in RStudio by going to the Packages pane, search up **stringr**, and then clicking on "User guides, package vignettes and other documentation". Or you can just access the web version [here](https://stringr.tidyverse.org/articles/regular-expressions.html)

There are several pre-built expressions that search for particular kinds of strings. Some useful ones are 

+ `[:alnum:]` : letters and numbers
+ `[:alpha:]` : letters
+ `[:digits:]`: digits
+ `[:blank:]` : spaces and tabs   

The `+` at the end means, grab one or more consecutive things of this kind. So that code looks for a contiguous set of letters and numbers starting from the beginning, and stops when it encounters the first non-letter or non-number. If you go back up, you'll realize that this code will stop just at the `(`, which isn't a letter or number.

**Exercise:** Modify the code above to change the search expression from `[:alnum:]+` to something else that will grab the same part of the string.

The hints give a few possibilities

```{r g1-hint-1}
mutate(mpg, trans = str_extract(trans, '[a-z]+'))
```

```{r g1-hint-2}
mutate(mpg, trans = str_extract(trans,'[:alpha:]+'))
```

```{r g1-hint-3}
mutate(mpg, trans = str_extract(trans, 'auto|manual'))
```


## Fitting 

```{r}
mpg1 <- mpg %>% 
  mutate(year = as.factor(year)) %>% 
  mutate(trans = str_extract(trans, '[:alnum:]+')) %>% 
  mutate(cyl = as.factor(cyl))
fit <- lm(cty ~ year + trans + cyl + drv + class, data=mpg1)
```

# Looking at the output

```{r}
library(broom)
out <- tidy(fit, conf.int = TRUE)

out
```

We need to fix the `terms` first. The factor variables have the unfortunate default behavior of being represened as <variable><level> with no real formatting. Also we could change the base level of `drv` to `f` (front wheel drive) from `4` (4-wheel drive), but we'll let it be for now.

# Munging output

We're going to build up the code to get this output data where we'd like it.

First, we need to identify the factor variables so that we can change them. Fortunately, R has already stored this information in the `fit` object.

```{r}
fit$xlevels
```

So we can see exactly which variables in the model are categorical and what their levels are. 

## Formatting the output

We're going to do a simple transformation: just add a `=` sign between the variable name and level for each categorical variable. Here's one way of approaching this.

```{r eq, exercise=TRUE}

for (n in names(fit$xlevels)){
  out <- out %>% 
    mutate(term = ifelse(str_detect(term, n), 
                         str_replace(term, n, paste(n,'= ')),
                         term))
}

```

Comment out parts of this code and really understand what it's doing. The end 
result gives us

```{r}
for (n in names(fit$xlevels)){
  out <- out %>% 
    mutate(term = ifelse(str_detect(term, n), 
                         str_replace(term, n, paste(n,'= ')),
                         term))
}

out
```

## Fixing the numbers

We're going to restrict the numbers to 2 decimals

```{r round, exercise=TRUE, exercise.eval=FALSE}
for(n in names(fit$xlevels)){
  out <- out %>% 
    mutate(term = ifelse(str_detect(term, n), 
                         str_replace(term, n, paste(n,'= ')),
                         term))
}

out <- out %>% 
  mutate_____(________, ~round(., 2))

```

```{r round-solution}
for(n in names(fit$xlevels)){
  out <- out %>% 
    mutate(term = ifelse(str_detect(term, n), 
                         str_replace(term, n, paste(n,'= ')),
                         term))
}

out <- out %>% 
  mutate_if(is.numeric, ~round(., 2))

```


This is quite reasonable for printing now.

```{r}
# for(n in names(fit$xlevels)){
#   out <- out %>% 
#     mutate(term = ifelse(str_detect(term, n), 
#                          str_replace(term, n, paste(n,'= ')),
#                          term))
# }
# 
# out <- out %>% 
#   mutate_if(is.numeric, ~round(., 2))

knitr::kable(out)
```


# Plotting

We're now in a position to plot this out

```{r final, exercise=TRUE, exercise.eval=TRUE}
# for(n in names(fit$xlevels)){
#   out <- out %>% 
#     mutate(term = ifelse(str_detect(term, n), 
#                          str_replace(term, n, paste(n,'= ')),
#                          term))
# }
# 
# out <- out %>% 
#   mutate_if(is.numeric, ~round(., 2))
# 
ggplot(out, aes(x = term, y = estimate, 
                ymin = conf.low, ymax = conf.high))+
  geom_pointrange() + 
  coord_flip()
```

Modify this code to make the plot prettier. The reference level should be 0



