---
title: "Slicing and filtering"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    css: css/style.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, comment = "")
library(tidyverse)

if(!('palmerpenguins' %in% installed.packages())){
  remotes::install_github('allisonhorst/palmerpenguins')
}
library(palmerpenguins)
```

## Introduction

The tidyverse provides several ways to programmatically subset rows of a 
`data.frame`/`tibble` object, based on either position or data values. The 
main two functions in the repertoire are `slice` and `filter`. Both are amenable
to pipes, since they take the tibble as the first argument.

Let's start with the _penguins_ data set. This set can be obtained by installing
the **palmerpenguins** package using `remotes::install_github('allisonhorst/palmerpenguins')`. 
This dataset had data on the characteristics of three penguin species. 

```{r Slicing-1}
glimpse(penguins, width=50)
```

This dataset is pre-loaded in this tutorial. 

## Slicing

### By position

Slicing by position is the same operation as `penguins[____,]`, where we would put some set of indices, like
`c(1,5,6,21)` or `1:10`. However, the `slice` function is pipe-friendly. 

```{r Slicing-2, exercise=TRUE}
penguins %>% slice(1, 2, 5, 8)
```

**Exercise:** Modify the code above to explore how to do different slices by index. You can combine indices like
`1, 3, 5:8, 12:15` in the same function call. Also note that you don't need the `c()` function to put these together
as you would for the `penguins[____,]` syntax.

### Slicing parts of the tibble

We might be familiar with the `head` and `tail` functions, which extract the first few and last few lines of 
the tibble. The functions `slice_head` and `slice_tail` do the same things, but do allow it to worked on grouped data as well. 

```{r Slicing-3}
penguins %>% slice_head(n = 5)
```

```{r Slicing-4, exercise=TRUE}
penguins %>% 
  group_by(species) %>% 
  slice_head(n=2)
```

These functions can also extract a fixed **proportion* of rows.

```{r Slicing-5, exercise=TRUE}
penguins %>% 
  group_by(species) %>% 
  slice_head(prop=0.02)
```

> The proportions are converted and rounded down to row numbers internally. So if you have 15 rows and ask for 
10% of the rows, you will get 1 row. 

### Sampling rows

Sampling rows is very useful either when you have a really large data and want to work on a random
subset first to develop your code, before applying it to the full data, or for sampling-based algorithms
like the bootstrap. 

```{r Slicing-6, exercise=TRUE}
set.seed(294)
penguins %>% 
  slice_sample(n = 10, replace = FALSE)
```

You can also sample by group, which can be useful when you want to preserve a certain structure in the sampled
data. 

```{r Slicing-7, exercise = TRUE}
set.seed(294)
penguins %>% 
  group_by(species) %>% 
  slice_sample(n = 2, replace = FALSE)
```

### Slice by value 

You can also slice a tibble based on minimum and maximum values of a variable.

```{r Slicing-8, exercise = TRUE}
penguins %>% 
  slice_max(body_mass_g, n = 4)
```

**Exercise:** Play around to see how grouping data by species or island before slicing changes the results.




