---
title: "Data cleaning: An example using WHO data"
output: 
  learnr::tutorial:
    css: css/style.css
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  This tutorial provides a walkthrough for tidying the WHO dataset on TB prevalence provided in the `tidyr` package. It showcases the capabilities of various `dplyr` single-table functions
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, message=F, warning=F)
library(tidyverse)
library(broom)

```

## The data

A subset of data from the World Health Organization (WHO) Global Tuberculosis Report
is included in the **tidyr** package. This data comprises counts of new TB cases recorded by group, for different countries and years. You can look at the details of this data by
typing `help(tidyr::who)` in the R console (you may need to open a separate instance of RStudio to do this,
if you're running RStudio with version less than 1.3)

Let's take a peek at this data


**About the data**

```{r, echo=FALSE}
data(who)
who %>% summarise_all(class) %>% 
  pivot_longer(names_to="Variable", values_to='Type', cols = everything())
```

**Top 3 rows**

```{r, echo=FALSE}
head(who, 3)
```

We can see two things immediately

1. There are columns that contain single pieces of information (number of TB cases) but the group information is in the column names. It is coded as
    + `new_`, followed by
    + method of diagnosis (`rel`,`sn`, `sp`, or `ep`), followed by
    + gender (`f` or `m`), followed by 
    + age group (`014`, `1525`, etc)
1. There is a naming inconsistency, where, for the `rel` variables, we have `newrel_...` instead of `new_rel_...`

## First, we fix names

If the naming pattern is consistent, it will help us with using `separate` later. So let's fix this now.

```{r, echo=TRUE}
names(who) <- str_replace(names(who), 'newrel','new_rel')
names(who) # These are he column names
```

> Yes, we just changed something in an in-built data set. You'll see that the data set now shows up in 
the Environment pane, since it's now local to our session. Don't worry, you haven't destroyed the original data. You can access it either by removing the local copy (`rm(who)`, or using the Environment pane), or directly accessing
the original data by `tidyr::who`

## Next we pivot

The first step in tidying the data is to create 3 columns for method of diagnosis, gender and age group, corresponding to each country and year. 

1. Identifier columns are `country`, `iso2` and `iso3`
1. The rest are columns we want to pivot

```{r pivot, exercise=TRUE}
cleaned_data <- who %>% 
  pivot_longer(names_to = "variable", values_to = "counts", cols = ____________)
```

<div id="pivot-hint">
There are a couple of ways you could specify the columns. One is using the `_____:_____` 
notation. The other takes advantage of the fact that each of those variables starts with `new_`, 
and so we could use `starts_with("new_")`
</div>

> Think about the different ways you can do this and how that might inform your own choice 
of variable names when you create your own data. You may be able to decide on a naming 
pattern that is both informative and makes data munging easier

## Separating columns

We have to separate the `variable` column into 3 columns: `diagnosis`, `sex` and `age`. 
One challenge to this is that the separators are different for the three parts.

### Method 1: Successive `separate`'s

My first attempt
was to do this in two steps

```{r sep1, exercise = TRUE}
cleaned_data <- who %>% 
  pivot_longer(names_to = 'variable', values_to = 'counts', starts_with('new')) %>% 
  separate(variable, into = c('dummy', 'diagnosis', 'gender_age'), sep = ____) %>% 
  separate(gender_age, into = c('gender', 'age'), sep = ____)
```

> Remember that the pipe takes the output at the previous step and makes it the input at
the next step. So, after the first `separate`, we have a separate variable for diagnosis, and 
so at the second `separate`, we only have to separate out gender and age groups

> The different possibilities for `sep` are available in the documentation for `separate`

```{r sep1-solution}
cleaned_data <- who %>% 
  pivot_longer(names_to = 'variable', values_to = 'counts', starts_with('new')) %>% 
  separate(variable, into = c('dummy', 'diagnosis', 'gender_age'), sep = '_') %>% 
  separate(gender_age, into = c('gender', 'age'), sep = 1)
```

### Regular expressions

Regular expressions can provide a powerful tool in extracting information from 
complex string patterns. 

>Learning regular expressions is beyond the scope of this class, 
but a starting point in working with regular expressions might be [here](https://medium.com/factory-mind/regex-tutorial-a-simple-cheatsheet-by-examples-649dc1c3f285) or [here](https://github.com/rstudio/cheatsheets/raw/master/regex.pdf). The R documentation
for regular expressions is also quite good (`?regex`).

So what is the pattern for the elements in `variable`? 

```
new_{diagnosis}_{gender}{age group}
```

where we know that 

+ diagnosis is a set of lower case letters
+ gender is `m` or `f`
+ age group is a set of numbers

For each of these, there is a regular expression that would match them (there are multiple choices, I'm putting down the simplest one to me).

Variable    | Nature       | Regular expression | Explanation
----------- | ------------ | ------------------ | -----------
diagnosis    | lower case letters | `[:lower:]+` | One or more consecutive lower case letters (a-z)
gender | single character (m/f) | `[mf]` | one of m and f
age_group | set of numbers  | `[:digit:]+` | one or more consecutive digits (0-9)


### Method 2: `extract`

We can use the function `tidyr::extract` to use regular expressions as a way to 
separate the columns of `variable`. It utilizes the regular expression concept of groups of
patterns, demarcated by `()`, to figure out what the separate pieces are. This idea is 
very much like what we wrote earlier for the pattern of entries in `variable`.

```{r, echo=TRUE}
cleaned_data <- who %>% 
  pivot_longer(cols=starts_with('new'), names_to = 'variable', values_to = 'counts') %>% 
  # gather(key = variable, value = counts, starts_with('new')) %>% 
  extract(variable, into = c('diagnosis','gender','age'),
          regex = 'new_([a-z]+)_([mf])([0-9]+)')
head(cleaned_data)
```

Pretty cool!!!

### Method 3: `pivot_longer`

The `pivot_longer` function in **tidyr** is quite flexible, and allows you to 
combine both the `gather` and `extract` steps in one go.

```{r, echo=TRUE}
cleaned_data <- who %>% 
  pivot_longer(cols = starts_with("new"),
               names_to = c('diagnosis','gender','age'),
               names_prefix = 'new_', # this gets removed first, so don't include in regex
               names_pattern = '([a-z]+)_([mf])([0-9]+)',
               values_to = 'counts')
head(cleaned_data)
```



