---
title: "Exploring ggplot2"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  This is a walkthrough to play with different aspects of ggplot2
---

```{r setup, include=FALSE}
library(learnr)
library(readr)
library(ggplot2)
library(mlbench)
library(patchwork)
data("PimaIndiansDiabetes2")
knitr::opts_chunk$set(echo = FALSE, warning=F, message=F, error=T)
beaches <- read_csv(system.file('extdata','beaches.csv', package='BIOF439Tutorials'))
```

## Loading a data package

You had installed and loaded the **mlbench** package into R as part of Formative 1 of 
week 2. Load that package into your current R session.

```{r}
pacman::p_load(mlbench)
data("PimaIndiansDiabetes2")
```


```{r load-package, exercise=TRUE}

```

To load a dataset from this package, you need to use a function called `data`. Use
this package to load the `PimaIndiansDiabetes2` dataset into R. You can then 
see the first few lines of the dataset using the `head` function.

```{r load-data, exercise=T}
data(___)
head(___)
```

## Univariate exploration (discrete variable)

Let's first look at the distribution of the number of pregancies in this dataset

```{r}
ggplot(PimaIndiansDiabetes2, aes(x = pregnant)) + 
  geom_bar(aes(fill = diabetes))
```

This is a bar plot that is displaying information from the `pregnant` column
of the `PimaIndiansDiabetes2` dataset, with colors based on values in the `diabetes` column. 
Let's try and recreate this plot. 

### Create a basic bar plot


```{r barplot, exercise = TRUE}
library(____)

ggplot(data = ___,
       mapping = aes(
         x = ___
       )) +
  geom_____()

```

```{r barplot-solution}
library(ggplot2) # or library(tidyverse)

ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = pregnant
       )) +
  geom_bar()
```


This should just be a bar plot with a single color. We're not displaying the diabetes status yet.

### Add code to make the plot above, with separate (default) colors by diabetic status.

```{r barplot2, exercise = TRUE}
ggplot(data = ___,
       mapping = aes(
         x = ___
       )) +
  geom_____(_________)
```

```{r barplot2-hint-1}
ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = pregnant
       )) +
  geom_bar(aes(_____))
```


```{r barplot2-hint-2}
ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = pregnant
       )) +
  geom_bar(aes(fill = diabetes))
```


### Exercise: barplot options

The `geom_bar` function has a `position` option to change how the bars for the
groups are positioned relative to each other, e.g., `geom_bar(position = ____)`. 
(**Hint:** Look at the documentation for `geom_bar` by typing `?geom_bar` in the console)

```{r barplot-quiz}
quiz(
  question('What are possible options for `position`?',
           answer('dodge', correct=TRUE),
           answer('next'),
           answer('fill', correct=TRUE),
           answer('stack', correct=TRUE),
           answer('default'),
           answer('side')
))
```

## Univariate exploration (continuous variable)

```{r}
plt1 <- ggplot(PimaIndiansDiabetes2, aes(x = mass, fill = diabetes))+
  geom_density(alpha = 0.3)
plt2 <- ggplot(PimaIndiansDiabetes2, aes(x = mass, color = diabetes))+
  geom_histogram(fill = 'white')
plt2 + plt1
```

We will develop two kinds of visualizations: histograms and density plots using
the `mass` variable from `PimaIndiansDiabetes2`.

### Creating a histogram

A histogram is just another possible geometry in **ggplot2** that is applied to 
continuous variables. Let's start by creating a histogram of `mass`. If you have
trouble with the particular geometry function, look in the package documentation
for possible functions. 

```{r hist1, exercise = TRUE}
ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = _____
       )) + 
  geom_______()
```

```{r hist1-solution}
ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = mass
       )) + 
  geom_histogram()
```

### Separate by group

We'll now create two histograms on the same panel, one for diabetics, and one
for non-diabetics

```{r hist2, exercise = TRUE}
ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = mass
       )) +
  geom_histogram(____(color = ____))
```

```{r hist2-hint-1}
ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = mass
       )) +
  geom_histogram(____(color = diabetes))
```

```{r hist2-hint-2}
ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = mass
       )) +
  geom_histogram(aes(color = diabetes))
```

### Customize appearance

Let's change the background of the histogram to white so we can see things better.

```{r hist3, exercise=TRUE}
ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = mass
       )) +
  geom_histogram(aes(color = diabetes), ______ = 'white')
```

```{r hist3-solution}
ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = mass
       )) +
  geom_histogram(aes(color = diabetes), fill = 'white')
```

### Exercise

```{r quiz2}
quiz(
  question("What aspect of a geometry does the option `color` control?",
           answer("The border", correct=TRUE),
           answer("The inside"),
           answer("The background of the plot")
     ))
```

> If a geometry doesn't have separate border and inside parts, like a line, the option `color` is used

### Convert this to a density plot

Eseentially the same code works to create a density plot, just changing the geometry. 
Modify the following code to create density plots, keeping the same options.

```{r dens, exercise=TRUE}
ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = mass
       )) +
  geom_histogram(aes(fill = diabetes))
```

```{r dens-solution}
ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = mass
       )) +
  geom_density(aes(fill = diabetes))
```


This hides one plot under the other, since, like a painter, it overlays one on the other. 

### Transparency

We can help "see through" the plots by adding an optional `alpha` value to the geometry. This 
takes values from 0 (completely transparent) to 1 (completely opaque). Play around with the 
`alpha` values to see how things change

```{r dens2, exercise = TRUE, exercise.eval=TRUE}
ggplot(data = PimaIndiansDiabetes2,
       mapping = aes(
         x = mass
       )) +
  geom_density(aes(fill = diabetes), alpha = 0.7)
```


## Bivariate exploration

We'll look at the distribution of women's ages by the number of pregnancies they have had. 

### Create a boxplot 

```{r bp1, exercise=TRUE, exercise.eval=TRUE}
ggplot(data = PimaIndiansDiabetes2, 
       mapping = aes(
         x = pregnant,
         y = age
       )) + 
  geom_boxplot()
```

Ooh!!! This is not what we wanted!!

### Create a boxplot, 2nd attempt

It turns out that the `pregnant` variable is of type `numeric`, while what we 
need is a `factor` type, which represents categorical variables. We can do this 
conversion on the fly.

```{r, echo=TRUE}
ggplot(data = PimaIndiansDiabetes2, 
       mapping = aes(
         x = as.factor(pregnant),
         y = age
       )) + 
  geom_boxplot()
```

Also, instead of those outlier points, let's take the whisker all the 
way to the end, so that it represents the maximum value. 

```{r, echo = TRUE}
ggplot(data = PimaIndiansDiabetes2, 
       mapping = aes(
         x = as.factor(pregnant),
         y = age
       )) + 
  geom_boxplot(coef=NULL) # Changed here
```

### Comparing groups

We've already seen how to create groups with different colors on a single pane. 

Modify the code below so we have side-by-side boxplots of different colors for
diabetics and non-diabetics

```{r bp3, exercise=TRUE, exercise.eval=TRUE}
ggplot(data = PimaIndiansDiabetes2, 
       mapping = aes(
         x = as.factor(pregnant),
         y = age
       )) + 
  geom_boxplot(coef=NULL) 
```

```{r bp3-solution}
ggplot(data = PimaIndiansDiabetes2, 
       mapping = aes(
         x = as.factor(pregnant),
         y = age
       )) + 
  geom_boxplot(aes(fill = diabetes), coef=NULL)
```

### Cleaning up

Let's modify the axis labels to English rather than code, since we 
modified a variable on-the-fly.

```{r bp4, exercise=TRUE}
ggplot(data = PimaIndiansDiabetes2, 
       mapping = aes(
         x = as.factor(pregnant),
         y = age
       )) + 
  geom_boxplot(aes(fill = diabetes), coef=NULL) + 
  labs(x = ______, y = _______, main = _______)
```

### Exercise

```{r q3}
quiz(
  question("If you have two continuous variables to compare, what geometries could you use?",
           answer('geom_bar'),
           answer('geom_point', correct=TRUE),
           answer('geom_line', correct=TRUE),
           answer('geom_violin'),
           answer('geom_col'),
           answer('geom_smooth', correct=TRUE))
)
```


## Facets

We saw with our histograms that layering histograms on top of each other doesn't
really work, especially if we have more than two groups.

We can split the groups up into facets to see each better. Let's explore
the age distributions by number of pregnancies, using density plots. 

First, let's see how the one-panel solution looks, even with transparency

```{r, echo=TRUE}
ggplot(PimaIndiansDiabetes2, 
       aes(x = age, fill = as.factor(pregnant)))+
  geom_density(alpha = 0.3)
```

> Note, I'm  not using `data=` and `mapping=` here. R functions can take arguments 
by position, and I know (you know too, now) that the first argument of `ggplot` is `data` and 
the second argument is `mapping`. So I can omit those and just write them out. However, 
if you want to put in function arguments out of order, you must name them using `data=` or whatever
the corresponding argument is.

### Splitting out by group

We can explode this plot out into facets using `facet_wrap`.

```{r fac, exercise=TRUE}
ggplot(PimaIndiansDiabetes2,
       aes(x = age))+  # Note, no more `fill`
  geom_density() + 
  facet_wrap(~as.factor(pregnant))
```

Change the above code to limit the grid to 4 columns

```{r fac-solution}
ggplot(PimaIndiansDiabetes2,
       aes(x = age))+  # Note, no more `fill`
  geom_density() + 
  facet_wrap(~as.factor(pregnant), ncol=4)
```

### Faceting by multiple variables

The `facet_grid` function provides much more control over faceting behavior. 

```{r grid, exercise = TRUE, exercise.eval=TRUE}
ggplot(mtcars, aes(x = mpg, y = wt)) + geom_point() + 
  facet_grid(am ~ gear)
```

Play around with this code by 

1. replacing `am` with `.`
1. replacing `gear` with `.`

```{r grid-hint-1}
ggplot(mtcars, aes(x = mpg, y = wt)) + geom_point() + 
  facet_grid(. ~ gear)
```


```{r grid-hint-2}
ggplot(mtcars, aes(x = mpg, y = wt)) + geom_point() + 
  facet_grid(am ~ .)
```

### The beaches data

The beaches data is already pre-loaded in this tutorial. 

Write code below that would create a facetted plot of the temperature distribution, 
with rows being years and columns being seasons. It should look something like this:

```{r}
ggplot(beaches, aes(x = temperature)) + 
  geom_density()+
  facet_grid(year ~ season_name)
```

```{r beaches, exercise = TRUE, exercise.lines = 8}

```

```{r beaches-solution}
ggplot(beaches, aes(x = temperature)) + 
  geom_density()+
  facet_grid(year ~ season_name)
```

### Fancy plotting

```{r fancy, exercise = TRUE, exercise.eval=TRUE}
ggplot(mpg, aes(drv, model)) +
  geom_point() +
  facet_grid(manufacturer ~ ., scales = "free", space = "free") +
  theme(strip.text.y = element_text(angle = 0))
```

Evaluate each option on your own, using the help file for `facet_grid`, 
to see how it affects the plot




